---     
title: "BayesSpace"
author: "Edward Zhao, Matt Stone, Xing Ren, and Raphael Gottardo"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{BayesSpace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width="100%",
  fig.width=7, fig.height=5,
  dpi=300, 
  # fig.path="figures/BayesSpace-",
  message=FALSE, warning=FALSE, error=FALSE
)
```

```{r setup}
library(BayesSpace)
library(ggplot2)
library(SingleCellExperiment)
library(tidyr)
```

## Preparing your experiment for BayesSpace

### Loading data

BayesSpace supports three ways of loading a `SingleCellExperiment` for analysis.
<!-- SE PRESENTAN TRES TECNICAS -->
<!-- PARECE QUE LA SEGUNDA ES LA MAS ADECUADA PARA NOSOTROS -->

<!-- Visium datasets processed with [Space -->
<!-- Ranger](https://support.10xgenomics.com/spatial-gene-expression/software/pipelines/latest/what-is-space-ranger) -->
<!-- can be loaded directly via the `readVisium()` function. This function takes only -->
<!-- the path to the Space Ranger output directory (containing the `spatial/` and -->
<!-- `filtered_feature_bc_matrix/` subdirectories) and returns a -->
<!-- `SingleCellExperiment`. -->

```{r readVisium, eval=FALSE}
# sce <- readVisium("path/to/spaceranger/outs/")
```

Second, all datasets analyzed for the BayesSpace manuscript are readily
accessible via the `getRDS()` function. This function takes two arguments - the
name of the dataset, and the name of the sample in the dataset.

```{r download}
melanoma <- getRDS(dataset = "2018_thrane_melanoma", sample = "ST_mel1_rep2")
# En este caso recurrimos directamente al repositorio local de BiocFileCache.
# Que hacer si queremos trabajar con un banco de datos propio.
# Tercera y última opción: Crear nosotros mismos nuestro SingleCellExperiment.
```

```{r}
melanoma_int_elementMetadata <- melanoma@int_elementMetadata
# Formal class DFrame
melanoma_int_elementMetadata@rownames
melanoma_int_elementMetadata@nrows # Contamos con 16148 genes
melanoma_int_elementMetadata@elementType
melanoma_int_elementMetadata@elementMetadata
melanoma_int_elementMetadata@metadata
melanoma_int_elementMetadata@listData

melanoma_int_colData <- melanoma@int_colData
# Formal class DFrame
melanoma_int_colData@rownames
melanoma_int_colData@nrows # Contamos con 293 spots (localizaciones)
melanoma_int_colData@elementType
melanoma_int_colData@elementMetadata
melanoma_int_colData@metadata
melanoma_int_colData_reducedDims <- as.data.frame(melanoma_int_colData@listData$reducedDims)
# Análisis de componentes principales para los 293 spots. ¿Por qué?
# Estaría bien saber cómo se realiza la construcción de este apartado
melanoma_int_colData@listData$altExps

melanoma_int_metadata <- melanoma@int_metadata
# List
melanoma_int_metadata$version
melanoma_int_metadata$spike_names
melanoma_int_metadata$size_factor_names
melanoma_int_metadata$scater

melanoma_rowRanges <- melanoma@rowRanges
# Large CompressedGRangesList
melanoma_rowRanges@unlistData
melanoma_rowRanges@unlistData@seqnames@values
melanoma_rowRanges@unlistData@seqnames@lengths
melanoma_rowRanges@unlistData@seqnames@elementMetadata
melanoma_rowRanges@unlistData@seqnames@metadata
melanoma_rowRanges@unlistData@ranges@start
melanoma_rowRanges@unlistData@ranges@width
melanoma_rowRanges@unlistData@ranges@NAMES
melanoma_rowRanges@unlistData@ranges@elementType
melanoma_rowRanges@unlistData@ranges@elementMetadata
melanoma_rowRanges@unlistData@ranges@metadata
melanoma_rowRanges@unlistData@strand@values
melanoma_rowRanges@unlistData@strand@lengths
melanoma_rowRanges@unlistData@strand@elementMetadata
melanoma_rowRanges@unlistData@strand@metadata
melanoma_rowRanges@unlistData@seqinfo@seqnames
melanoma_rowRanges@unlistData@seqinfo@seqlengths
melanoma_rowRanges@unlistData@seqinfo@is_circular
melanoma_rowRanges@unlistData@seqinfo@genome
melanoma_rowRanges@unlistData@elementMetadata@rownames
melanoma_rowRanges@unlistData@elementMetadata@nrows
melanoma_rowRanges@unlistData@elementMetadata@elementType
melanoma_rowRanges@unlistData@elementMetadata@elementMetadata
melanoma_rowRanges@unlistData@elementMetadata@metadata
melanoma_rowRanges@unlistData@elementMetadata@listData
melanoma_rowRanges@unlistData@elementType
melanoma_rowRanges@unlistData@metadata
melanoma_rowRanges@elementMetadata@rownames
melanoma_rowRanges@elementMetadata@nrows
melanoma_rowRanges@elementMetadata@elementType
melanoma_rowRanges@elementMetadata@elementMetadata
melanoma_rowRanges@elementMetadata@metadata
melanoma_rowRanges@elementMetadata@listData$gene_id
# Identificador de los 16148 genes
melanoma_rowRanges@elementMetadata@listData$gene_name
# Nombre de los 16148 genes
table(melanoma_rowRanges@elementMetadata@listData$is.HVG)
# True/False
# Recoge si nuestro gen es HVGs: Highly Variable Genes
melanoma_rowRanges@elementType
melanoma_rowRanges@metadata
melanoma_rowRanges@partitioning@end
melanoma_rowRanges@partitioning@NAMES
melanoma_rowRanges@partitioning@elementType
melanoma_rowRanges@partitioning@elementMetadata
melanoma_rowRanges@partitioning@metadata

melanoma_colData <- melanoma@colData
# Formal class DFrame
melanoma_colData@rownames
# Posicion de los 293 spots: 7x15 . . . 27x22
melanoma_colData@nrows
melanoma_colData@elementType
melanoma_colData@elementMetadata
melanoma_colData@metadata
melanoma_colData@listData$row
# Fila para cada uno los 293 spots
melanoma_colData@listData$col
# Columna para cada uno de los 293 spots
melanoma_colData@listData$sizeFactor
# Factor de elevación. ¿De dónde sale?

melanoma_assays <- melanoma@assays
# Large SimpleAssays
melanoma_assays_counts <- melanoma_assays@data@listData$counts
# LargedgCMatrix
melanoma_assays_counts@i
# Son 715865 entradas.
# Cada entrada hace referencia a un determinado gen
# 16148 genes x 293 spots = 4731364
# No se recoge información de todos los genes en todas las localizaciones
which(melanoma_assays_counts@i == 1)
# El gen 1 solo está para 45/293 localizaciones
which(melanoma_assays_counts@i == 2)
# El gen 2 solo está para 292/293 localizaciones
which(melanoma_assays_counts@i == 3)
# El gen 3 solo está para 119/293 localizaciones
# Debemos descubrir cuales son las localizaciones de nuestros genes
# Por ejemplo, para el gen 1, cuales son los 45 spots
localizacion_final_spot <- c()
a <- 0
auxiliar <- 1
count <- 1
spot <- c()
for(i in 1:length(melanoma_assays_counts@i)){
  print(i)
  if(a <= melanoma_assays_counts@i[i]){
    a <- melanoma_assays_counts@i[i]
  } else{
    localizacion_final_spot <- c(localizacion_final_spot, i - 1)
    spot <- c(spot, rep(count, i - auxiliar))
    a <- 0
    auxiliar <- i
    count <- count + 1
  }
}
spot <- c(spot, rep(293, i + 1 - auxiliar))
# Podemos hacer un data.frame con el gen y su correspondiente spot
melanoma_assays_counts@p[-1]
# La informacion que recogemos en melanoma_assays_counts@p y localizacion_final_spot
# Es la misma. Imbeciles
melanoma_assays_counts@Dim
melanoma_assays_counts@Dimnames
melanoma_assays_counts@x
# Recuento de cada uno de los genes para cada uno de los distintos spots
melanoma_assays_counts@factors
# El procedimiento con base logaritmica seria similar
melanoma_assays@data@elementType
melanoma_assays@data@elementMetadata
melanoma_assays@data@metadata
datos <- data.frame(gen = melanoma_assays_counts@i + 1, 
                    recuento = melanoma_assays_counts@x,
                    spot = spot)
# Modificacion de datos (formato amigable)
datos <- spread(datos, key = gen, value = recuento)
datos[is.na(datos)] <- 0

melanoma_elementMetadata <- melanoma@elementMetadata
# Formal class DFrame
melanoma_elementMetadata@rownames
melanoma_elementMetadata@nrows
melanoma_elementMetadata@elementType
melanoma_elementMetadata@elementMetadata
melanoma_elementMetadata@metadata
melanoma_elementMetadata@elementMetadata

melanoma_metadata <- melanoma@metadata
# List
melanoma_metadata$sample
melanoma_metadata$dataset
melanoma_metadata$BayesSpace.data
```


Finally, `SingleCellExperiment` objects can be constructed manually from a
counts matrix and tables of row and column data. BayesSpace only requires that
spot array coordinates be provided as columns named `row` and `col` in
`colData`. (Note that enhancement of Visium datasets additionally requires the
pixel coordinates of each spot in the tissue image, but in this case the dataset
should be loaded with `readVisium()`, which loads these data automatically.)

```{r manual.sce, eval=FALSE}
library(Matrix)

# rowData <- read.csv("path/to/rowData.csv", stringsAsFactors=FALSE)
# colData <- read.csv("path/to/colData.csv", stringsAsFactors=FALSE, row.names=1)
# counts <- read.csv("path/to/counts.csv.gz",
#                    row.names=1, check.names=F, stringsAsFactors=FALSE))
# 
# sce <- SingleCellExperiment(assays=list(counts=as(counts, "dgCMatrix")),
#                             rowData=rowData,
#                             colData=colData)

##### ----- PLANTEAMOS EL EJEMPLO ASOCIADO A LA AYUDA DE LA FUNCIÓN ----- #####
# Con el planteamiento aquí propuesto trabariamos con:
# 200 genes y 100 spots
# Una construccion interesante si plantearemos trabajar con datos propios

ncells <- 100
u <- matrix(rpois(20000, 5), ncol = ncells)
v <- log2(u + 1)

pca <- matrix(runif(ncells*5), ncells) # Estas son las instrucciones relativas a la PCA
tsne <- matrix(rnorm(ncells*2), ncells) # Valorar objetivamente su construccion.

sce <- SingleCellExperiment(assays = list(counts = u, logcounts = v),
                            reducedDims = SimpleList(PCA = pca, tSNE = tsne))
# Construccion similar a la planteada en la opcion anterior
se <- SummarizedExperiment(assays = list(counts = u, logcounts = v))
# Experimento más sencillo
##### ----- PLANTEAMOS EL EJEMPLO ASOCIADO A LA AYUDA DE LA FUNCIÓN ----- #####
```

We'll continue with the melanoma sample from the 2018 Spatial Transcriptomics
paper for the remaining examples in this vignette.

### Pre-processing data

BayesSpace requires minimal data pre-processing, but we provide a helper
function to automate it.

`spatialPreprocess()` log-normalizes the count matrix and performs PCA on the
top `n.HVGs` highly variable genes, keeping the top `n.PCs` principal
components. Additionally, the spatial sequencing platform is added as metadata
in the `SingleCellExperiment` for downstream analyses. If you do not wish to
rerun PCA, running `spatialPreprocess()` with the flag `skip.PCA=TRUE` will only
add the metadata BayesSpace requires.

Here, we omit log-normalization as all datasets available through `getRDS()`
already include log-normalized counts.

```{r preprocess}
set.seed(102)
melanoma2 <- spatialPreprocess(melanoma, platform = "ST", 
                               n.PCs = 7, n.HVGs = 2000, log.normalize = FALSE)
# Recordar del analisis anterior del tipo de objeto
# Teniamos 2000 genes que era HVGs
# Ademas contabamos con 7 componenentes para nuestro PCA

# Comentarios al respecto
# Variable platform
# Nos permite definir una estructura hexagonal o cuadrada como en el articulo
# ¿Valoramos posibles relaciones de vencidad que se podrian estudiar?

# En este caso melanoma1 y melanoma2 son identicos
# Pero y si modificamos n.PCs y n.HVGs
melanoma2 <- spatialPreprocess(melanoma, platform = "ST", 
                               n.PCs = 15, n.HVGs = 4000, log.normalize = FALSE)
# Ahora melanoma dos presenta distintas caracteristicas en las dos variables que hemos modificado
# ¿Estudiar cual es el numero optimo de PCs y HVGs?
```

## Clustering

### Selecting the number of clusters

We can use the `qTune()` and `qPlot()` functions to help choose `q`, the number
of clusters to use in our analysis.
<!-- CONOCEMOS UNA GRAN VARIEDAD DE TÉCNICAS DESDE MINERÍA DE DATOS. EXPLORAR -->

* `qTune()` runs the BayesSpace clustering algorithm for multiple specified
values of `q` (by default, 3 through 7) and computes their average
pseudo-log-likelihood. It accepts any arguments to `spatialCluster()`.
* `qPlot()` plots the pseudo-log-likelihood as a function of `q`; we suggest
choosing a `q` around the elbow of this plot.

```{r tuning_q}
# Parece que la técnica de clusterización va muy orientada al tipo de objeto.
# Observar la construcción de la función qTune para poder adaptarla a otros intereses.
melanoma3 <- qTune(melanoma, qs = seq(2, 10), platform = "ST", d = 7)
# Es el mismo SingleCellExperiment que antes
# No tiene absolutamente nada nuevo
qPlot(melanoma3)
# Parece asequible la funcion y su modificacion para nuestros conocimientos de clustering
```

### Clustering with BayesSpace

METER COMPONENTE ESPACIAL EN EL PRIMER CLUSTER 


The `spatialCluster()` function clusters the spots, and adds the predicted
cluster labels to the `SingleCellExperiment`. Typically, as we did for the
analyses in the paper, we suggest running with at least 10,000 iterations
(`nrep=10000`), but we use 1,000 iteration in this demonstration for the sake of
runtime. (Note that a random seed must be set in order for the results to be
reproducible.)

```{r cluster}
# Mismo comentario que en el chunk anterior
set.seed(149)
melanoma4 <- spatialCluster(melanoma, q = 4, platform = "ST", d = 7,
                           init.method = "mclust", model = "t", gamma = 2,
                           nrep = 1000, burn.in = 100,
                           save.chain = TRUE)
# ¿Queremos extraer el maximo jugo de esta funcion y la anterior? Yo creo que si
# Tratar de acercar la funcion aqui definida a nuestros conocimientos de EstadisticaBayesiana
# melanoma4 si que tiene ya diferencias con respecto a melanoma. 
# Aparecen el cluster inicial y el asignado (necesario para nuestra representacion grafica)
```

Both the mclust initialization (`cluster.init`) and the BayesSpace cluster
assignments (`spatial.cluster`) are now available in the SingleCellExperiment's
`colData`.

```{r cluster.results}
head(colData(melanoma))
```

### Visualizing spatial clusters

We can plot the cluster assignments over the spatial locations of the spots with
`clusterPlot()`.

```{r cluster.plot, fig.width=7, fig.height=5}
clusterPlot(melanoma4)
# Función construida mediante el trabajo con ggplot2
# La idea final debe ser encontrar una asociacion entre los clusters establecidos
# Una relacion con la cantidad de genes presentes en esos clusters
# Este debe ser el paso final para sacar las conclusiones oportunas
```

As `clusterPlot()` returns a `ggplot` object, it can be customized by composing
with familiar `ggplot2` functions. Additionally, the argument `palette` sets the
colors used for each cluster, and `clusterPlot()` takes additional arguments to
`geom_polygon()` such as `size` or `color` to control the aesthetics of the spot
borders.

```{r cluster.plot.customize, fig.width=7, fig.height=5}
clusterPlot(melanoma4, palette = c("purple", "red", "blue", "yellow"), color = "black") +
  theme_bw() +
  xlab("Column") +
  ylab("Row") +
  labs(fill = "BayesSpace\ncluster", title = "Spatial clustering of ST_mel1_rep2")
```


## Enhanced resolution

### Clustering at enhanced resolution

The `spatialEnhance()` function will enhance the resolution of the principal
components, and add these PCs as well as predicted cluster labels at subspot
resolution to a new `SingleCellExperiment`. As with our demonstration of
`spatialCluster()` above, we are using fewer iterations for the purpose of this
example (`nrep=1000`) than we recommend in practice (`nrep=100000` or greater).
Note that the `jitter_scale` parameter should be tuned so that proposals for 
updating subspot-level expression are accepted around 30% of the time. This can
be evaluated using `mcmcChain(melanoma.enhanced, "Ychange")`, where the chain 
should stabilize to 0.25-0.40. Typically 1000-2500 iterations are sufficient to 
evaluate if `jitter_scale` should be increased if acceptance is too high or 
decreased if acceptance is too low. After tuning, proceed to a full run of 
`spatialEnhance` with more iterations.

```{r enhance, eval=TRUE}
melanoma.enhanced <- spatialEnhance(melanoma4, q = 4, platform = "ST", d = 7,
                                    model = "t", gamma = 2,
                                    jitter_prior = 0.3, jitter_scale = 3.5,
                                    nrep = 1000, burn.in = 100,
                                    save.chain = TRUE)
# Esta es posiblemente la funcion mas importante
# Nos plantea la construccion de subspots
# Pasamos de 293 a 2637
# Podriamos volver a analizar el objeto como en el primer apartado
# Sacar todo su jugo en cuanto a los datos
```

The enhanced `SingleCellExperiment` includes an index to the parent spot in the
original `sce` (`spot.idx`), along with an index to the subspot. It adds the
offsets to the original spot coordinates, and provides the enhanced cluster
label (`spatial.cluster`).

```{r enhance.results}
head(colData(melanoma.enhanced))
```

We can plot the enhanced cluster assignments as above.

```{r enhance.plot, eval=TRUE, fig.width=7, fig.height=5}
clusterPlot(melanoma.enhanced)
# Mi duda ahora mismo:
# Conociamos con los datos originales el recuento de cada gen para cada spot
# ¿Como sabemos ahora el recuento de los distintos genes para diferentes spots?
# No sabemos
```

### Enhancing the resolution of gene expression

<!-- BayesSpace operates on the principal components of the gene expression matrix, -->
<!-- and `spatialEnhance()` therefore computes enhanced resolution PC vectors. -->
<!-- Enhanced gene expression is not computed directly, and is instead imputed using -->
<!-- a regression algorithm. For each gene, a model using the PC vectors of each spot -->
<!-- is trained to predict the spot-level gene expression, and the fitted model is -->
<!-- used to predict subspot expression from the subspot PCs.  -->
<!-- YO CREO QUE ESTA ES LA PRINCIPAL PARTE EN LA QUE DEBEMOS TRABAJAR -->
<!-- ANALIZAR / MODIFICAR LA REGRESIÓN PROPUESTA Y TRABAJAR EN ESA LÍNEA -->

Gene expression enhancement is implemented in the `enhanceFeatures()` function.
BayesSpace predicts expression with
[`xgboost`](https://xgboost.readthedocs.io/en/latest/) by default, but linear
and Dirichlet regression are also available via the `model` argument. When using
`xgboost`, we suggest automatically tuning the `nrounds` parameter by setting it
to 0, although this comes at the cost of increased runtime (~4x slower than a
pre-specified `nrounds` in practice).

`enhanceFeatures()` can be used to impute subspot-level expression for all
genes, or for a subset of genes of interest. Here, we'll demonstrate by
enhancing the expression of four marker genes: PMEL (melanoma), CD2 (T-cells),
CD19 (B-cells), and COL1A1 (fibroblasts). 

```{r enhanceFeatures}
markers <- c("PMEL", "CD2", "CD19", "COL1A1")
melanoma.enhanced <- enhanceFeatures(melanoma.enhanced, melanoma4,
                                     feature_names=markers,
                                     nrounds=0)
```

By default, log-normalized expression (`logcounts(sce)`) is imputed, although
other assays or arbitrary feature matrices can be specified.

```{r enhanced.logcount}
logcounts(melanoma.enhanced)[markers, 1:5]
```

Diagnostic measures from each predictive model, such as `rmse` when using
`xgboost`, are added to the `rowData` of the enhanced dataset.

```{r enhanced.rmse}
rowData(melanoma.enhanced)[markers, ]
```


### Visualizing enhanced gene expression

Spatial gene expression is visualized with `featurePlot()`.

```{r enhanced.featurePlot}
featurePlot(melanoma.enhanced, "PMEL")
# En este sentido. ¿Qué conclusion debemos sacar con respecto a la clusterizacion inicial?
# Que se busca comparando las 2 graficas
```

Here, we compare the spatial expression of the imputed marker genes.

```{r enhanced.markers, fig.width=12, fig.height=8}
enhanced.plots <- purrr::map(markers, function(x) featurePlot(melanoma.enhanced, x))
patchwork::wrap_plots(enhanced.plots, ncol=2)
```

And we can compare to the spot-level expression.

```{r compare.resolution, fig.width=16, fig.height=8}
spot.plots <- purrr::map(markers, function(x) featurePlot(melanoma, x))
patchwork::wrap_plots(c(enhanced.plots, spot.plots), ncol=4)
```


## Accessing Markov chains

If `save.chain` is set to `TRUE` in either `spatialCluster()` or
`spatialEnhance()`, the chain associated with the respective MCMC run is
preserved to disk as an HDF5 file. The path to this file is stored in the
SingleCellExperiment's metadata at `metadata(sce)$h5.chain`, and can be read
directly using `mcmcChain()`.

The chain is provided as a `coda::mcmc` object, which can be analyzed with
[TidyBayes](https://mjskay.github.io/tidybayes/) or as a matrix. The object has
one row per iteration, with the values of the parameters concatenated across the
row. Columns are named with the parameter name and index (if any).

```{r mcmcChain, eval=TRUE}
chain <- mcmcChain(melanoma4)
chain[1:5, 1:5]
medias <- apply(chain, 2, mean)
View(cbind(colnames(chain), medias))
```

To remove the HDF5 file from disk and remove its path from the metadata, use
`removeChain()`.
